Comparison of several stochastic and deterministic derivative-free global optimization algorithms\thanks{The study was supported by the Russian Science Foundation, project No 16-11-10150.}

### 2019-02-13 2019/MOTOR_Sovrasov/motor_paper.tex
MOTOR: update paper

%%%%%ДОБАВИЛ ДЛЯ РУССКОГО ТЕКСТА
\usepackage[utf8x]{inputenc}
\usepackage[english,russian]{babel}
\usepackage{cmap}

@@

\begin{abstract}
In this paper popular sequential open-source solvers are compared against
Globalizer solver, which is developed at the Lobachevsky State University.
The Globalizer is designed to solve problems with black-box objectives satisfying the Lipschitz condition and shows
competitive performance with other similar solvers. The comparison is done on several sets of
challenging multi-extremal benchmark functions. Also this work considers a method of
heuristic hyperparameters control for Globalizer allowing to reduce amount of
initial tuning before optimization. The proposed scheme allows rapidly increase convergence speed
of Globalizer by switching between ``local'' and ``global'' search phases in runtime.
% We would like to encourage you to list your keywords within
% the abstract section using the \keywords{...} command.
\keywords{deterministic global optimization $\cdot$ stochastic global optimization
  $\cdot$ parallel numerical methods $\cdot$ derivative-free algorithms $\cdot$ black-box
optimization}
\end{abstract}
%
\section{Introduction}
\begin{Russian}

Задача отысканя глобального минимума нелинейных невыпуклых функций традиционно считается одной из самых трудных
задач математического программирования. Зачастую она оказывается сложнее, чем локальная оптимизация в существенно многомерном пространстве.
Для последней может оказаться достаточно применения простейшего метода градиентного спуска или pattern search algorithms \cite{torczon1997},
в то время как чтобы \textit{гаранитрованно} отыскать глобальный оптимум методам
оптимизации приходится накапливать информацию о поведении целевой функции во всей области поиска \cite{Jones2009,Paulavicius2011,Evtushenko2013,strSergGO}. В последнее время стали популярны
различные стохастические алгоритмы глобально оптимизации, прежде всего эволюционные \cite{Storn1997, SCHLUTER2009, KennedyEberhart1995}. Они имеют довольно простую структуру, позволяют решать задачи большой размерности, однако обеспечивают глобальную сходимость только в вероятностном смысле.

В данной работе рассмотрены open-source реализации восьми различных методов глобальной оптимизации, представленные
в библиотеке NLOpt\cite{nlopt} и пакете SciPY\cite{scipy}.
Все алгоритмы были протестированы на наборе из 900 существенно многоэкстремальных функций, который был сгенерирован с
помощью специализированных генераторов задач \cite{Gaviano2003, grishaginClass}.
\end{Russian}

@@

\begin{Russian}
Параметр $r$ из (\ref{eq:step2}) непосредственно влияет на глобальную сходимость AGS (see \cite{strSergGO}, Chapter 8):
при достаточно большом значении $r$ метод гарантированно сходится ко всем глобальным минимумам целевой функции.
В то же время, согласно (\ref{eq:step3_2}), при бесконечно большом значении $r$ AGS превращается в полный перебор
по равномерной сетке.

В идеальном случае для обеспечения наибольшей скорости сходимости оценка константы Липшица из (\ref{eq:step2})
не должна быть слишком завышенной, но на практике реальное значение $L$ из (\ref{eq:lip}) неизвестно и приходится
либо брать заведомо большое значение $r$, либо проводить несколько запусков AGS с разными параметрами. Чтобы в некотрой
степени решить обозначенную проблему выбора $r$ будем использовать следующую схему:
\begin{itemize}
  \item совершить $q$ итераций AGS при $r=r_{max}$;
  \item совершить $q$ итераций AGS при $r=r_{min}$;
  \item повторять предыдущие шаги до схоимости или исчерпания лимита итараций.
\end{itemize}

В указанном алгоритме $r_{min} < r_{max}$, $q > 1$. Вместо одного параметра $r$ теперь
необходимо выбирать 3, однако, согласно численным экспериментам, сделать это проще, чем найти оптимальное значение $r$.
Интуитивно практическую эффективность предложенной схемы можно объяснить тем, что теперь работа метода происходит в
двух режимах: глобальный поиск при $r=r_{max}$ и локальный при $r=r_{min}$. Если во время фазы глобального посика
метод приблизился к глобальному минимуму, то во время следующей фазы оценка глобального минимума будет быстро уточнена.
Если двух фаз не хватило, то процесс продолжается. Таким образом выбирается более оптимальный trade-off between
exploration and exploitation. Далее будем обозначать метод, использующий описанную схему как AGS-AR.
\end{Russian}

@@

Let us suppose a test problem to be solved if the optimization method executes the scheduled trial
\(y^k\) in a \(\delta\)-vicinity of the global minimum \(y^*\), i.e. $\left\|y^k-y^*\right\|\leq \delta
= 0.01\left\|b-a\right\|$, where \(a\) and \(b\) are the left and the right boundaries of the hypercube
from (\ref{eq:task}). If this relation is not fulfilled before the expiration of the limit of the number of
trials, the problem was considered to be unsolved. The limit of the number of trials was set
for each problem class according to the dimensionality and complexity (see Table \ref{tab:limits}).

\begin{table}
\begin{center}
\caption{Trials limits for the test problem classes}
  \begin{tabular}{|l|{c}|}
    \hline
  Problems class & Trials limit\\
  \hline
  \(F_{GR}\) & 5000 \\
  \hline
  GKLS 2d Simple & 8000 \\
  \hline
  GKLS 2d Hard & 9000 \\
  \hline
  GKLS 3d Simple & 15000 \\
  \hline
  GKLS 3d Hard & 25000 \\
  \hline
  GKLS 4d Simple & 150000 \\
  \hline
  GKLS 4d Hard & 250000 \\
  \hline
  GKLS 5d Simple & 350000 \\
  \hline
  GKLS 5d Hard & 600000 \\
  \hline
  \end{tabular}
  \label{tab:limits}
\end{center}
\end{table}

@@

The results of various algorithms on different problem classes depend on the adjustments of
algorithms directly. In most cases, the authors of software implementations are oriented onto the
problems of medium difficulty. In order to obtain a satisfactory result when solving the essentially
multiextremal problems, a correction of some parameters is required. When conducting the
comparison, the following parameters for the methods were employed:
\begin{itemize}
  \item in the AGS-AR method, the parameter of alternation the
  global and local stages $q$ was set to be equal to $100\cdot N^2$, also $r_{min}=3,\:r_{max}=2\cdot r_{min}$;
  \item in the DIRECT and DIRECT\(l\) methods, the parameter \(\epsilon=10^{-4}\);
  \item in the SDA method, the parameter \(visit=2.72\).
\end{itemize}

@@

\begin{table}
\begin{center}
\caption{Averaged number of trials executed by optimization methods for solving the test
optimization problems}
\resizebox{\textwidth}{!}{%
  \begin{tabular}{|l|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|}
    \hline
    & AGS & AGS-AR & CRS & DIRECT & DIRECT\(l\) & MLSL & SDA & DE & StoGO \\
  \hline
  \(F_{GR}\)     & 193.1 &  \textbf{158.3} & 400.3 & 182.3 & 214.9 & 947.2 & 691.2 & 1257.3 &
1336.8 \\
  \hline
  GKLS 2d Simple &  254.9 & 217.6 & 510.6 & \textbf{189.0} & 255.2 & 556.8 & 356.3 & 952.2
& 1251.5 \\
  \hline
  GKLS 2d Hard   &  728.7 & \textbf{488.0} & 844.7 & 985.4 & 1126.7 & 1042.5 & 1637.9 &
1041.1 & 2532.2 \\
  \hline
  GKLS 3d Simple &  1372.1 & 1195.3 & 4145.8 & \textbf{973.6} & 1477.8 & 4609.2 & 2706.5 &
5956.94 & 3856.1 \\
  \hline
  GKLS 3d Hard   &  3636.1 & \textbf{1930.5} & 6787.0 & 2298.7 & 3553.3 & 5640.1 & 4708.4 &
6914.3 & 7843.2 \\
  \hline
  GKLS 4d Simple &  26654.1 & 11095.7 & 37436.8 & \textbf{7824.3} & 15994.1 & 41514.3 &
21417.9 & 19157.7 & 59895.4 \\
  \hline
  GKLS 4d Hard   &  54536.8 &  \textbf{23167.8} &  73779.3 &  23204.4 &  54489.9 &  80247.2
&  68815.5 &  27466.1 &  109328.1  \\
  \hline
  GKLS 5d Simple &  29810.0 & 11529.0 & 143575.0 & \textbf{7166.5} & 13970.5 & 52647.6 &
34255.3 & 73074.5 & 91580.4 \\
  \hline
  GKLS 5d Hard   &  113129.1 & 67652.7 & 165192.8 & \textbf{66327.4} & 164390.6 & 138766.2
& 116973.1 & 105496.9 & 155123.8 \\
  \hline
\end{tabular}}
  \label{tab:trials}
\end{center}
\end{table}

\begin{table}
\begin{center}
\caption{Number of test optimization problems solved by methods}
  \begin{tabular}{|l|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|}
    \hline
    & AGS & AGS-AR & CRS & DIRECT & DIRECT\(l\) & MLSL & SDA & DE & StoGO \\
  \hline
  \(F_{GR}\)     &  100 & 100 & 76  & 100 & 100 & 97  & 96  & 96  & 67\\
  \hline
  GKLS 2d Simple &  100 & 100 & 85  & 100 & 100 & 100 & 100 & 98  & 90\\
  \hline
  GKLS 2d Hard   &  100 & 97 & 74  & 100 & 100 & 100 & 93  & 85  & 77 \\
  \hline
  GKLS 3d Simple &  100 & 100  & 75  & 100 & 100 & 100 & 89  & 86  & 44 \\
  \hline
  GKLS 3d Hard   &  100  & 100   & 72   & 100  & 99   & 100  & 88   & 77   & 43 \\
  \hline
  GKLS 4d Simple &  100 & 100 & 46  & 100 & 100 & 94  & 78  & 59  & 16 \\
  \hline
  GKLS 4d Hard   &  100 & 100 & 47  & 99  & 97  & 94  & 72  & 32  & 10  \\
  \hline
  GKLS 5d Simple &  100 & 100 & 68  & 100 & 100 & 98  & 100 & 77  & 9  \\
  \hline
  GKLS 5d Hard   &  97  & 100  & 42  & 100 & 90  & 79  & 84  & 48  & 8 \\
  \hline
  \end{tabular}
  \label{tab:solved}
\end{center}
\end{table}

>>>>

@@

\begin{abstract}
In this paper popular sequential open-source solvers are compared against
Globalizer solver, which is developed at the Lobachevsky State University.
The Globalizer is designed to solve problems with black-box objectives satisfying the Lipschitz condition and shows
competitive performance with other similar solvers. The comparison is done on several sets of
challenging multi-extremal benchmark functions. Also this work considers a method of
heuristic hyperparameters control for Globalizer allowing to reduce amount of
initial tuning before optimization. The proposed scheme allows rapidly increase convergence speed
of Globalizer by switching between ``local'' and ``global'' search phases in runtime.
% We would like to encourage you to list your keywords within
% the abstract section using the \keywords{...} command.
\keywords{deterministic global optimization $\cdot$ stochastic global optimization
  $\cdot$ parallel numerical methods $\cdot$ derivative-free algorithms $\cdot$ black-box
optimization}
\end{abstract}
%
\section{Introduction}

The problem of finding the global minima of the nonlinear nonconvex functions is considered to be
one of the most difficult mathematical programming problems traditionally. Often, it appears to be
more complex than the local optimization in an essentially multidimensional space.
For the latter, the application of the simplest gradient descent method or of the pattern search
algorithms may appear to be sufficient \cite{torczon1997}
whereas in order to \textit{guarantee} the finding of the global optimum, the optimization methods
have to
accumulate the information on the behavior of the objective function in the whole search domain
\cite{Jones2009,Paulavicius2011,Evtushenko2013,strSergGO}.
Recently, various stochastic global optimization algorithms, first of all, the evolution ones
\cite{Storn1997, SCHLUTER2009, KennedyEberhart1995} became popular. These ones have rather
simple structure and allow solving the problems of large dimensionality. However, these methods
provide the global convergence in the probabilistic meaning only.

In the present work, the open-source implementations of the eight different global optimization
methods included into the NLOpt library \cite{nlopt} and the SciPY package \cite{scipy} are
considered.
All algorithms were tested on a set of 900 essentially multiextremal functions, which has been
generated with the use of special problem generators \cite{Gaviano2003, grishaginClass}.

@@

The parameter $r$ from (\ref{eq:step2}) affects the global convergence of AGS directly (see
\cite{strSergGO}, Chapter 8):
at high enough value of $r$, the method converges to all global minima of the objective function with
guarantee.
At the same time, according to (\ref{eq:step3_2}), at the infinitely high value of $r$, AGS turns into
the brute force search method on a uniform grid.

In the ideal case, in order to provide the highest convergence speed, the estimate of the Lipschitz
constant from (\ref{eq:step2})
should not be too overestimated, but in practice the actual value of $L$ from (\ref{eq:lip}) in
unknown, and one has either to take an obviously overestimated value of $r$ or to execute several
runs of AGS with different parameters. In order to resolve the problem of choosing $r$ to some extent,
let us use the following scheme:
\begin{itemize}
  \item execute $q$ iterations of AGS with $r=r_{max}$;
  \item execute $q$ iterations of AGS with $r=r_{min}$;
  \item repeat the above steps either until convergence or until the allowed number of iterations are
exhausted.
\end{itemize}

In the above algorithm, $r_{min} < r_{max}$, $q > 1$. Instead one parameter $r$, now
3 ones should be selected. However, according to the results of the numerical experiments, it is easier
than to find the optimal value of $r$.
Intuitively, the practical efficiency of the proposed scheme can be explained by the fact that now the
operation of the method takes place in two modes: the global search with $r=r_{max}$ and the local
one with $r=r_{min}$. If during the global search phase, the method approached the global minimum
whereas during the next phase, the estimate of the global minimum  would be refined rapidly.
If two phases are not enough, the process is continued. This way, a better trade-off
between the exploration and the exploitation is achieved.
Further, we will denote the method utilizing the scheme described above as AGS-AR.

@@

Let us suppose a test problem to be solved if the optimization method executes the scheduled trial
\(y^k\) in a \(\delta\)-vicinity of the global minimum \(y^*\), i.e. $\left\|y^k-y^*\right\|\leq \delta
= \alpha\left\|b-a\right\|$, where \(a\) and \(b\) are the left and the right boundaries of the hypercube
from (\ref{eq:task}), $\alpha$ is relative precision. If this relation is not fulfilled before the expiration of the limit of the number of
trials, the problem was considered to be unsolved. The limit of the number of trials and $\alpha$ were set
for each problem class according to the dimensionality and complexity (see Table \ref{tab:limits}).

\begin{table}
\begin{center}
\caption{Trials limits and relative precision for the test problem classes}
  \begin{tabular}{|l|{c}|{c}|}
    \hline
  Problems class & Trials limit & $\alpha$\\
  \hline
  \(F_{GR}\) & 5000 & 0.01 \\
  \hline
  GKLS 2d Simple & 8000 & 0.01 \\
  \hline
  GKLS 2d Hard & 9000 & 0.01 \\
  \hline
  GKLS 3d Simple & 15000 & 0.01 \\
  \hline
  GKLS 3d Hard & 25000 & 0.01 \\
  \hline
  GKLS 4d Simple & 150000 & $\sqrt[4]{10^{-6}}$ \\
  \hline
  GKLS 4d Hard & 250000 & $\sqrt[4]{10^{-6}}$ \\
  \hline
  GKLS 5d Simple & 350000 & $\sqrt[5]{10^{-7}}$ \\
  \hline
  GKLS 5d Hard & 600000 & $\sqrt[5]{10^{-7}}$ \\
  \hline
  \end{tabular}
  \label{tab:limits}
\end{center}
\end{table}

@@

The results of various algorithms on different problem classes depend on the adjustments of
algorithms directly. In most cases, the authors of software implementations are oriented onto the
problems of medium difficulty. In order to obtain a satisfactory result when solving the essentially
multiextremal problems, a correction of some parameters is required. When conducting the
comparison, the following parameters for the methods were employed:
\begin{itemize}
  \item in the AGS-AR method, the parameter of alternation the
  global and local stages $q$ was set to be equal to $50\cdot\log_2(N-1)\cdot N^2$, also $r_{min}=3,\:r_{max}=2\cdot r_{min}$;
  \item in the DIRECT and DIRECT\(l\) methods, the parameter \(\epsilon=10^{-4}\);
  \item in the SDA method, the parameter \(visit=2.72\).
\end{itemize}

@@

\begin{table}
\begin{center}
\caption{Averaged number of trials executed by optimization methods for solving the test
optimization problems}
\resizebox{\textwidth}{!}{%
  \begin{tabular}{|l|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|}
    \hline
    & AGS & AGS-AR & CRS & DIRECT & DIRECT\(l\) & MLSL & SDA & DE & StoGO \\
  \hline
  \(F_{GR}\)     & 193.1 & 248.3 & 400.3 & \textbf{182.2} & 214.9 & 947.2 & 691.2 & 1257.3 & 1336.8 \\
  \hline
  GKLS 2d Simple & 254.9 & 221.6 & 510.6 & \textbf{189.0} & 255.2 & 556.8 & 356.3 & 952.2 & 1251.5 \\
  \hline
  GKLS 2d Hard   & \textbf{728.7} & 785.0 & 844.7 & 985.4 & 1126.7 & 1042.5 & 1637.9 & 1041.1 & 2532.2 \\
  \hline
  GKLS 3d Simple &  1372.1 & 1169.5 & 4145.8 & \textbf{973.6} & 1477.8 & 4609.2 & 2706.5 & 5956.9 & 3856.1 \\
  \hline
  GKLS 3d Hard   &  3636.1 & \textbf{1952.1} & 6787.0 & 2298.7 & 3553.3 & 5640.1 & 4708.4 & 6914.3 & 7843.2 \\
  \hline
  GKLS 4d Simple &  5729.8 & \textbf{4919.1} & 19883.6 & 7328.8 & 15010.0 & 41484.8 & 22066.0 & 6271.2 & 29359.2 \\
  \hline
  GKLS 4d Hard   &  13113.4 & \textbf{12860.1} & 27137.4 & 22884.4 & 55596.1 & 80220.1 & 68048.0 & 12487.6 & 58925.5  \\
  \hline
  GKLS 5d Simple &  \textbf{5821.5} & 6241.3 & 62921.7 & 5966.1 & 10795.5 & 52609.2 & 34208.8 & 20859.4 & 69206.8 \\
  \hline
  GKLS 5d Hard   &  \textbf{17008.6} & 21555.1 & 87563.9 & 61657.3 & 148637.8 & 138011.8 & 115634.6 & 26850.0 & 141886.5 \\
  \hline
\end{tabular}}
  \label{tab:trials}
\end{center}
\end{table}

\begin{table}
\begin{center}
\caption{Number of test optimization problems solved by methods}
  \begin{tabular}{|l|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|}
    \hline
    & AGS & AGS-AR & CRS & DIRECT & DIRECT\(l\) & MLSL & SDA & DE & StoGO \\
  \hline
  \(F_{GR}\)     &  100 & 100 & 76 & 100 & 100 & 97 & 96 & 96 & 67\\
  \hline
  GKLS 2d Simple &  100 & 100 & 85 & 100 & 100 & 100 & 100 & 98 & 90\\
  \hline
  GKLS 2d Hard   &  100 & 97 & 74 & 100 & 100 & 100 & 93 & 85 & 77 \\
  \hline
  GKLS 3d Simple &  100 & 100 & 75 & 100 & 100 & 100 & 89 & 86 & 44 \\
  \hline
  GKLS 3d Hard   &  100 & 100 & 72 & 100 & 99 & 100 & 88 & 77 & 43 \\
  \hline
  GKLS 4d Simple &  100 & 100 & 74 & 100 & 100 & 94 & 82 & 68 & 72 \\
  \hline
  GKLS 4d Hard   &  100 & 100 & 60 & 99 & 99 & 94 & 73 & 55 & 69  \\
  \hline
  GKLS 5d Simple &  100 & 100 & 86 & 100 & 100 & 98 & 100 & 88 & 82  \\
  \hline
  GKLS 5d Hard   &  100 & 100 & 77 & 100 & 93 & 79 & 86 & 77 & 78 \\
  \hline
  \end{tabular}
  \label{tab:solved}
\end{center}
\end{table}
